# Assessing the best practice for surrogate variable analysis in epigenome-wide association studies

```{r setoptions, eval = TRUE, echo = FALSE}
opts_chunk$set(echo = FALSE, 
            warning = FALSE, 
            message = FALSE, 
            fig.width = 12, 
            fig.height = 7,
            cache = FALSE,
            fig.align = 'center')
```

```{r call_source}
read_chunk("~/PACE/pace_ad/reports/pace_ad_report.r")
```

``` {r load_data, results = "hide", message = FALSE, warning = FALSE}
```

## Aim:
To test the length of time taken to generate surrogate variables (SVs) using various different parameters and compare the SVs generated

## Methods:

### Data
Used the DNA methylation data from ALSPAC mothers (timepoint = FOM, n = `r `). This is DNA methylation measured in whole blood samples. All CpGs were used for analysis. 

Phenotype data was simulated using the rnorm() function in R.

### Analysis
Parameters tested:
* SVA package - sva and smartsva
* Number of CpGs - increasing from 20,000 to 300,000 by increments of 20,000 + the full set of CpGs (`r`)
* 